---
title: "Simulate correlation between average expression and treatment-specificity"
output: html_notebook
---

# Load packages
```{r}
rm(list = ls())
gc()
library(matrixStats) # to take row medians
library(ggplot2)
library(tidyverse)
```

# SIMULATION PARAMETERS
```{r}
nGenes = 1000 # number of genes
nTreatments = 10 # number of treatments/environments
nExperimentsPerTreatment = 5 # number of experiments per treatment
nSims = 1000 # number of simulations to run
binomP = 0.4 # prob parameter for binomial distribution, used to draw zeros
nbinomSize = 100 # size paramter for negative binomial, draws nonzero expression values
nbinomP = 0.1 # prob paramter for negative binomial, draws nonzero expression values
```

# SIMULATION ONE
## Assume that gene expression values are drawn from negative binomial
```{r}
### FUNCTIONS
# calcualte specificity index
specificity = function(x){
  sum(1-(x/max(x)))/(length(x)-1)
}

# Test specificity function
# specificity(c(0,0,0,0,0,0,0,0,10000))
# specificity(c(1,100,5,5,10))
# specificity(c(10,10,10,10,10,10))

# store output for simulations
sim1Cors = rep(NA, length.out = nSims)
# simAvgExps = vector(mode = "list", length = nSims)
# simSpecs = vector(mode = "list", length = nSims)

### RUN SIMULATION
for(j in 1:nSims){
  # Generate random expression matrix
  expMat = matrix(rnbinom(n = nGenes*nTreatments*nExperimentsPerTreatment, size = nbinomSize, p = nbinomP), nrow = nGenes, ncol = nTreatments*nExperimentsPerTreatment)

  # calculate average expression across treatments
  avgExp = rowMeans(expMat)

  # Matrix to store treatment profiles (expression averages within each treatment)
  treatmentProfiles = NULL
  # calculate specifcity index for each gene
  for(i in 1:nTreatments){
    # print(c(2*(i-1)+1,2*i))
    # Subset rows in groups of nTreatments, calculate mean of experiments from same treatment
    treatmentProfile = rowMeans(expMat[,c((nExperimentsPerTreatment*(i-1)+1):(nExperimentsPerTreatment*i))])
    # save experiment means to build treatment profiles for each gene
    treatmentProfiles = cbind(treatmentProfiles, treatmentProfile)
  }

  # calculate specificity from profiles
  simSpec = apply(treatmentProfiles, MARGIN = 1, FUN = specificity)

  # calculate correltion between average expression and specificity
  # simAvgExps[[j]]= avgExp
  # simSpecs[[j]] = simSpec
  sim1Cors[j] = cor(avgExp, simSpec)

}

### SIMULATION RESULTS
# hist(unlist(simAvgExps))
# summary(unlist(simAvgExps))

# hist(unlist(simSpecs))
# summary(unlist(simSpecs))

hist(sim1Cors, xlab = "Correlation between average expression and treatment specificity", main = "")
summary(sim1Cors)
```

# SIMULATION TWO
## Expression values are drawn from a different zero-inflated negative binomial
```{r}

# Zero-inflated negative binomial
# n = number of gene expression values to draw
# size = size parameter for negative binomial
# p2 = prob parameter for negative binomial
# p1 = prob paramter for binomial, used to draw zeros
rznbinom = function(n, size, p1, p2){
  # generate random vector of zeros and ones
  output = rbinom(n = n, size = 1, prob = p1)

  # For every element with a one, replace with value from negative binomial
  indices = which(output == 1)
  nOnes = length(indices)
  output[indices] = rnbinom(nOnes, size = 100, prob = p2)
  
  return(output)
}

# test rznbinom function
hist(rznbinom(1000, 100, 0.7, 0.1))

# store output for simulations
sim2Cors = rep(NA, length.out = nSims)

### RUN SIMULATION
for(j in 1:nSims){
  # Generate random expression matrix
  expMat = matrix(rznbinom(n = nGenes*nTreatments*nExperimentsPerTreatment, size = nbinomSize, p1 = binomP, p2 = nbinomP), nrow = nGenes, ncol = nTreatments*nExperimentsPerTreatment)

  # calculate average expression across treatments
  avgExp = rowMeans(expMat)

  # Matrix to store treatment profiles (expression averages within each treatment)
  treatmentProfiles = NULL
  # calculate specifcity index for each gene
  for(i in 1:nTreatments){
    # print(c(2*(i-1)+1,2*i))
    # Subset rows in groups of nTreatments, calculate mean of experiments from same treatment
    treatmentProfile = rowMeans(expMat[,c((nExperimentsPerTreatment*(i-1)+1):(nExperimentsPerTreatment*i))])
    # save experiment means to build treatment profiles for each gene
    treatmentProfiles = cbind(treatmentProfiles, treatmentProfile)
  }

  # calculate specificity from profiles
  simSpec = apply(treatmentProfiles, MARGIN = 1, FUN = specificity)

  # calculate correltion between average expression and specificity
  sim2Cors[j] = cor(avgExp, simSpec)

}

### SIMULATION RESULTS
hist(sim2Cors, main = "", xlab = "Correlation between average expression and treatment specificity")
summary(sim2Cors)
```

# SIMULATION THREE
## Try to remove correlation between average expression and treatment-specificity by looking at max expression instead of mean expression
```{r}

# store output for simulations
sim3Cors = rep(NA, length.out = nSims)

### RUN SIMULATION
for(j in 1:nSims){
  # Generate random expression matrix
  expMat = matrix(rznbinom(n = nGenes*nTreatments*nExperimentsPerTreatment, size = nbinomSize, p1 = binomP, p2 = nbinomP), nrow = nGenes, ncol = nTreatments*nExperimentsPerTreatment)

  # calculate average expression across treatments
  avgExp = rowMaxs(expMat)

  # Matrix to store treatment profiles (expression averages within each treatment)
  treatmentProfiles = NULL
  # calculate specifcity index for each gene
  for(i in 1:nTreatments){
    # print(c(2*(i-1)+1,2*i))
    # Subset rows in groups of nTreatments, calculate mean of experiments from same treatment
    treatmentProfile = rowMeans(expMat[,c((nExperimentsPerTreatment*(i-1)+1):(nExperimentsPerTreatment*i))])
    # save experiment means to build treatment profiles for each gene
    treatmentProfiles = cbind(treatmentProfiles, treatmentProfile)
  }

  # calculate specificity from profiles
  simSpec = apply(treatmentProfiles, MARGIN = 1, FUN = specificity)

  # calculate correltion between average expression and specificity
  sim3Cors[j] = cor(avgExp, simSpec, use = "pairwise.complete.obs")

}

### SIMULATION RESULTS
hist(sim3Cors)
summary(sim3Cors)
```


# SIMULATION FOUR
## Look at correlation between average expression (excluding zeros) and treatment-specificity
```{r}
# Function for average of vector, excluding zeros
meanExcludeZeros = function(x){
  mean(x[(x != 0)])
}

mean(0:10)
meanExcludeZeros(0:10)

# store output for simulations
sim4Cors = rep(NA, length.out = nSims)

### RUN SIMULATION
for(j in 1:nSims){
  # Generate random expression matrix
  expMat = matrix(rznbinom(n = nGenes*nTreatments*nExperimentsPerTreatment, size = nbinomSize, p1 = binomP, p2 = nbinomP), nrow = nGenes, ncol = nTreatments*nExperimentsPerTreatment)

  # calculate average expression across treatments
  avgExp = apply(expMat, MARGIN = 1, FUN = meanExcludeZeros)

  # Matrix to store treatment profiles (expression averages within each treatment)
  treatmentProfiles = NULL
  # calculate specifcity index for each gene
  for(i in 1:nTreatments){
    
    # Subset rows in groups of nTreatments, calculate mean of experiments from same treatment
    treatmentProfile = rowMeans(expMat[,c((nExperimentsPerTreatment*(i-1)+1):(nExperimentsPerTreatment*i))])
    # save experiment means to build treatment profiles for each gene
    treatmentProfiles = cbind(treatmentProfiles, treatmentProfile)
  }

  # calculate specificity from profiles
  simSpec = apply(treatmentProfiles, MARGIN = 1, FUN = specificity)

  # calculate correltion between average expression and specificity
  sim4Cors[j] = cor(avgExp, simSpec, use = "pairwise.complete.obs")

}

### SIMULATION RESULTS
hist(sim4Cors)
summary(sim4Cors)
```

# Visualize results of all sims side-by-side
```{r}

myColorPalette = c("#ff4141", "#f6ee8b", "#52624a", "#7b834a")

simData = data.frame(simulation = as.factor(c(rep("NegBinom, mean expression", times = length(sim1Cors)), rep("ZINegBinom, mean expression", times = length(sim2Cors)), rep("ZINegBinom, max expression", times = length(sim3Cors)), rep("ZINegBinom, mean expression, exclude zeros", times = length(sim4Cors)))), correlations = c(sim1Cors, sim2Cors, sim3Cors, sim4Cors))

ggplot(simData, aes(x=correlations, color = simulation)) + 
  theme_classic() +
  labs(x = "Correlation between expression level and treatment-specificity", y = "Density") +
  scale_color_manual(values=myColorPalette) +
  geom_density()
  # stat_density(aes(x=correlations, color = simulation), geom = "line")
```

# What happens to tau as number of treatments approaches infinity?
```{r}
# Specificity seems to not increase very rapidly as more treatments are added, potentially means convergence
small = replicate(100, specificity(rznbinom(1000,100,0.4,0.1)))
medium = replicate(100, specificity(rznbinom(10000,100,0.4,0.1)))
large = replicate(100, specificity(rznbinom(100000,100,0.4,0.1)))

summary(small)
summary(medium)
summary(large)

# Increasing probability of nonzero expression -> lowers tau
small = replicate(100, specificity(rznbinom(1000,100,0.9,0.1)))
medium = replicate(100, specificity(rznbinom(10000,100,0.9,0.1)))
large = replicate(100, specificity(rznbinom(100000,100,0.9,0.1)))

summary(small)
summary(medium)
summary(large)

# What happens when you manipulate the mean and variance of negbinom?
# Increase variance -> increase tau, seemingly very slowly
# increase size parameter -> variance increases linearly
# increase probability parameter -> variance increases hyperbolically
small = replicate(100, specificity(rznbinom(1000,1000,0.4,0.1)))
medium = replicate(100, specificity(rznbinom(10000,1000,0.4,0.1)))
large = replicate(100, specificity(rznbinom(100000,1000,0.4,0.1)))

summary(small)
summary(medium)
summary(large)

# GRID SEARCH, HOW DO ZINEGBINOM PARAMETERS AFFECT TAU
# Define grid boundry
p1List = seq(from = 0.1, to = 1, by = 0.1)
p2List = seq(from = 0.1, to = 1, by = 0.1)
nList = seq(from = 100, to = 1000, by = 100)

# grid search with three parameters: n, p1, p2
# Define a named list of parameter values
gs <- list(nValues = nList,
           p1Values = p1List,
           p2Values = p2List) %>% 
  cross_d() # Convert to data frame grid

gs = as.data.frame(gs)

# loop over grid points
tauResult = c()
for(i in 1:nrow(gs)){
  gsRow = gs[i,]
  nPoint = as.numeric(gsRow[1])
  p1Point = as.numeric(gsRow[2])
  p2Point = as.numeric(gsRow[3])

  tauResult = c(tauResult, specificity(rznbinom(1000,nPoint,p1Point,p2Point)))
  # tauResult = c(tauResult, mean(replicate(100, specificity(rznbinom(1000,nPoint,p1Point,p2Point)))))
}

gs$tau = tauResult

# Look at how these different factors influence tau
# Tau is overwhelmingly determined by the proportion of observations that are zeros
summary(lm(tau ~ nValues + p1Values + p2Values, gs))
summary(lm(tau ~ p1Values, gs))
summary(lm(tau ~ p2Values, gs))
summary(lm(tau ~ nValues, gs))
```
